---
title: "Housing Policies in Practice: A Dive into the Evidence"
subtitle: "Housing Policies in Practice"
author: "Hussain Hadah (he/him)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: ["default", "assets/tulane-fonts.css", "assets/tulane.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: true # show a title slide with YAML information
    includes:
      in_header: "assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
layout: true
<div style="position: absolute;left:20px;bottom:5px;color:black;font-size: 12px;">`r rmarkdown::metadata$author` (Tulane) | `r rmarkdown::metadata$subtitle` | `r format(Sys.time(), '%d %B %Y')`</div>

<!--- `r rmarkdown::metadata$subtitle` | `r format(Sys.time(), '%d %B %Y')`-->


---

```{r, load_refs, echo=FALSE, cache=FALSE, message=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'authoryear', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("assets/example.bib", check = FALSE)
top_icon = function(x) {
  icons::icon_style(
    icons::fontawesome(x),
    position = "fixed", top = 10, right = 10
  )
}
```

```{r setup, include=FALSE}
# xaringanExtra::use_scribble() ## Draw on slides. Requires dev version of xaringanExtra.

options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=4, fig.width=6,
  out.width="748px", out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
```


```{r, cache=FALSE, message=FALSE, warning=FALSE, include=TRUE, eval=TRUE, results=FALSE, echo=FALSE, tidy.opts = list(width.cutoff = 50), tidy = TRUE}
## Load and install the packages that we'll be using today
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tictoc, parallel, pbapply, future, future.apply, furrr, RhpcBLASctl, memoise, here, foreign, mfx, tidyverse, hrbrthemes, estimatr, ivreg, fixest, sandwich, lmtest, margins, vtable, broom, modelsummary, stargazer, fastDummies, recipes, dummy, gplots, haven, huxtable, kableExtra, gmodels, survey, gtsummary, data.table, tidyfast, dtplyr, microbenchmark, ggpubr, tibble, viridis, wesanderson, censReg, rstatix, srvyr, formatR, sysfonts, showtextdb, showtext, thematic, sampleSelection, RefManageR, DT, googleVis, png, grid)
# devtools::install_github("thomasp85/patchwork")
remotes::install_github("mitchelloharawild/icons")
remotes::install_github("ROpenSci/bibtex")

# devtools::install_github("ajdamico/lodown")
## My preferred ggplot2 plotting theme (optional)
## https://github.com/hrbrmstr/hrbrthemes
# scale_fill_ipsum()
# scale_color_ipsum()
font_add_google("Fira Sans", "firasans")
font_add_google("Fira Code", "firasans")

showtext_auto()

theme_customs <- theme(
  text = element_text(family = 'firasans', size = 16),
  plot.title.position = 'plot',
  plot.title = element_text(
    face = 'bold', 
    colour = thematic::okabe_ito(8)[6],
    margin = margin(t = 2, r = 0, b = 7, l = 0, unit = "mm")
  ),
)

colors <-  thematic::okabe_ito(5)

# theme_set(theme_minimal() + theme_customs)
theme_set(hrbrthemes::theme_ipsum() + theme_customs)
## Set master directory where all sub-directories are located
mdir <- "/Users/hhadah/Documents/GiT/urban-econ/slides/01-week1/tue"

## Set working directory

# COLOR PALLETES
library(paletteer) 
# paletteer_d("awtools::a_palette")
# paletteer_d("suffrager::CarolMan")

### COLOR BLIND PALLETES
#paletteer_d("colorblindr::OkabeIto")
#paletteer_d("colorblindr::OkabeIto_black")
# paletteer_d("colorBlindness::paletteMartin")
# paletteer_d("colorBlindness::Blue2DarkRed18Steps")
# paletteer_d("colorBlindness::SteppedSequential5Steps")
# paletteer_d("colorBlindness::PairedColor12Steps")
# paletteer_d("colorBlindness::ModifiedSpectralScheme11Steps")
colorBlindness <- paletteer_d("colorBlindness::Blue2Orange12Steps")
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# scale_colour_paletteer_d("colorBlindness::ModifiedSpectralScheme11Steps", dynamic = FALSE)
# To use for fills, add
  scale_fill_manual(values="colorBlindness::Blue2Orange12Steps")

# To use for line and point colors, add
  scale_colour_manual(values="colorBlindness::Blue2Orange12Steps")
  #<a><button>[Click me](#sources)</button></a>
```

```{css, echo=F}
    /* Table width = 100% max-width */

    .remark-slide table{
        width: auto !important; /* Adjusts table width */
    }

    /* Change the background color to white for shaded rows (even rows) */

    .remark-slide thead, .remark-slide tr:nth-child(2n) {
        background-color: white;
    }
    .remark-slide thead, .remark-slide tr:nth-child(n) {
        background-color: white;
    }
```

## Overview of the Evidence

### We will go over several papers that study the effect of housing policies on affordability 

### Will will cover the impact of:

1. Exclusionary Zoning
2. Homeowners Associations
3. Redlining
4. Upzoning

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Exclusionary Zoning

---
## Definition

### Exclusionary zoning refers to the practice of using zoning ordinances to exclude certain types of land uses from a given community

### It has been used to exclude low-income families and minorities from living in certain neighborhoods

### First laws were passed in 20th century to prevent the construction of affordable apartments

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Aaronson Et Al. (2020) & Aaronson Et Al. (2021)

---
## Overview

### What is the long-run effects of the 1930s Home Owners' Loan Corporation (HOLC) redlining maps on socioeconomic status and economic mobility?

--
### Redlining reduced home ownership rates, house values, and rents and increased racial segregation in later decades 

### People born in 1970s and 1980s were significantly affected by the HOLC maps

---
## Background: Redlining

.blockquote[
### `r icons::fontawesome("comment-dots")` Redlining 
Derived from the red shading that demarcated the lowest-ranked D neighborhoods in which the racial makeup of neighborhoods were explicit factors that were often pivotal in assigning grades.
]

---
## HOLC Maps

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture01.png"))
```

---
## Summary Statistics

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture02.png"))
```

---
## Effect of Redlining: D-C Border

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture03.png"))
```

---
## Effect of Redlining: C-B Border

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture04.png"))
```

---
## Effect of Redlining: Income & Upward Mobility

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture05.png"))
```

---
## Effect of Redlining: Poverty & Incarceration

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture06.png"))
```

---
## Effect of Redlining: Family Structure

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture07.png"))
```

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Clarke and Freedman (2019)

---
## What is the relationship between being in a homeowners association and housing prices?

### Having an HOA increases housing prices by 4.2% on average

### HOAs has a large effect on income and racial segregation

---
## Relationship between HOAs and Diversity

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture08.png"))
```

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Song (2021)

---
## How do zoning restrictions affect housing prices, construction, and segregation?

### Zoning restrictions increases:

#### A 100% increase in minimum lot size increases:

1. Housing prices by 14%
2. Rent by 6%
3. Intensity of segregation

---
## Minimum Lot Size and Segregation

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture09.png"))
```

---
## Minimum Lot Size and Housing Prices and Rent

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture10.png"))
```

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Jarvis (2021)

---
## What is the effect of NIMBYism on local planning?

#### What is NIYMBYism?

- Not In My Backyard (NIMBY) is a term used to describe opposition by residents to a proposal for a new development close to them

--
- The author finds strong local opposition to wind projects

-  Increased wind power deployment costs by 10-29%

--
- Similar opposition could and does apply to local housing projects

---
## Example of NIMBYism

- [Tree Murder Song](https://twitter.com/gramsofgnats/status/1227658713973940224?s=61&t=Ypr5tIPuoNPFVrHneOfZ8g)

- [Egg Lady](https://www.youtube.com/watch?v=ZilZ6P2P49Q)

- New apartments mean new housing but what about the poor [Zucchini](https://www.youtube.com/watch?v=3VnwT-dzJUs) in her backyard

- [The Falafel Gambit](https://medium.com/yimbyneoliberal/the-great-falafel-debacle-27231edce59e)

--
### So, is local opposition to projects democratic?

---

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/ruby.png"))
```

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Greenaway-McGrevy and  Phillips (2023)

---
## What is the impact of upzoning on construction?

- Upzoning is when single family zoning restrictions are loosened

### Upzoning stimulated construction

### Upzoning can stimulate housing supply

---
## Upzoning stimulates construction

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/pic11.png"))
```

---
## Upzoning stimulates construction

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/pic12.png"))
```

---
## Upzoning stimulates construction

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/pic13.png"))
```

---
## Upzoning stimulates construction

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/pic14.png"))
```

---
class: tulane-blue
background-image: url(assets/TulaneLogo-white.svg)
background-size: 260px
background-position: 5% 95%

# Tulane and the Murphy Institute are hosting a housing sympoisum

- Friday and Saturday April 12 and 13 


```{r gen_pdf, include = FALSE, cache = FALSE, eval = FALSE}
library(renderthis)
# https://hhadah.github.io/xaringan/index.html

to_pdf(from = "~/Documents/GiT/urban-econ/slides/01-week1/tue/01-class1.html", 
       to = "~/Documents/GiT/urban-econ/slides/01-week1/tue/01-class1.pdf")
to_pdf(from = "~/Documents/GiT/urban-econ/slides/01-week1/tue/01-class1.html")
```
