---
title: "Urban Economics"
subtitle: "Rent Control"
author: "Hussain Hadah (he/him)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: ["default", "assets/tulane-fonts.css", "assets/tulane.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: true # show a title slide with YAML information
    includes:
      in_header: "assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
layout: true
<div style="position: absolute;left:20px;bottom:5px;color:black;font-size: 12px;">`r rmarkdown::metadata$author` (Tulane) | `r rmarkdown::metadata$subtitle` | `r format(Sys.time(), '%d %B %Y')`</div>

<!--- `r rmarkdown::metadata$subtitle` | `r format(Sys.time(), '%d %B %Y')`-->

---
class: title-slide
background-image: url("assets/TulaneLogo-white.svg"), url("assets/title-image1.jpg")
background-position: 10% 90%, 100% 50%
background-size: 160px, 50% 100%
background-color: #0148A4
# .text-shadow[.white[Outline for Today]]

<ol>
    <li><h4 class="white"> </h4></li>
</ol>

---

```{r, load_refs, echo=FALSE, cache=FALSE, message=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'authoryear', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("assets/example.bib", check = FALSE)
top_icon = function(x) {
  icons::icon_style(
    icons::fontawesome(x),
    position = "fixed", top = 10, right = 10
  )
}
```

```{r setup, include=FALSE}
# xaringanExtra::use_scribble() ## Draw on slides. Requires dev version of xaringanExtra.

options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=4, fig.width=6,
  out.width="748px", out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
```


```{r, cache=FALSE, message=FALSE, warning=FALSE, include=TRUE, eval=TRUE, results=FALSE, echo=FALSE, tidy.opts = list(width.cutoff = 50), tidy = TRUE}
## Load and install the packages that we'll be using today
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tictoc, parallel, pbapply, future, future.apply, furrr, RhpcBLASctl, memoise, here, foreign, mfx, tidyverse, hrbrthemes, estimatr, ivreg, fixest, sandwich, lmtest, margins, vtable, broom, modelsummary, stargazer, fastDummies, recipes, dummy, gplots, haven, huxtable, kableExtra, gmodels, survey, gtsummary, data.table, tidyfast, dtplyr, microbenchmark, ggpubr, tibble, viridis, wesanderson, censReg, rstatix, srvyr, formatR, sysfonts, showtextdb, showtext, thematic, sampleSelection, RefManageR, DT, googleVis, png, grid)
# devtools::install_github("thomasp85/patchwork")
remotes::install_github("mitchelloharawild/icons")
remotes::install_github("ROpenSci/bibtex")

# devtools::install_github("ajdamico/lodown")
## My preferred ggplot2 plotting theme (optional)
## https://github.com/hrbrmstr/hrbrthemes
# scale_fill_ipsum()
# scale_color_ipsum()
font_add_google("Fira Sans", "firasans")
font_add_google("Fira Code", "firasans")

showtext_auto()

theme_customs <- theme(
  text = element_text(family = 'firasans', size = 16),
  plot.title.position = 'plot',
  plot.title = element_text(
    face = 'bold', 
    colour = thematic::okabe_ito(8)[6],
    margin = margin(t = 2, r = 0, b = 7, l = 0, unit = "mm")
  ),
)

colors <-  thematic::okabe_ito(5)

# theme_set(theme_minimal() + theme_customs)
theme_set(hrbrthemes::theme_ipsum() + theme_customs)
## Set master directory where all sub-directories are located
mdir <- "/Users/hhadah/Documents/GiT/urban-econ/slides/01-week1/tue"

## Set working directory

# COLOR PALLETES
library(paletteer) 
# paletteer_d("awtools::a_palette")
# paletteer_d("suffrager::CarolMan")

### COLOR BLIND PALLETES
#paletteer_d("colorblindr::OkabeIto")
#paletteer_d("colorblindr::OkabeIto_black")
# paletteer_d("colorBlindness::paletteMartin")
# paletteer_d("colorBlindness::Blue2DarkRed18Steps")
# paletteer_d("colorBlindness::SteppedSequential5Steps")
# paletteer_d("colorBlindness::PairedColor12Steps")
# paletteer_d("colorBlindness::ModifiedSpectralScheme11Steps")
colorBlindness <- paletteer_d("colorBlindness::Blue2Orange12Steps")
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# scale_colour_paletteer_d("colorBlindness::ModifiedSpectralScheme11Steps", dynamic = FALSE)
# To use for fills, add
  scale_fill_manual(values="colorBlindness::Blue2Orange12Steps")

# To use for line and point colors, add
  scale_colour_manual(values="colorBlindness::Blue2Orange12Steps")
  #<a><button>[Click me](#sources)</button></a>
```

```{css, echo=F}
    /* Table width = 100% max-width */

    .remark-slide table{
        width: auto !important; /* Adjusts table width */
    }

    /* Change the background color to white for shaded rows (even rows) */

    .remark-slide thead, .remark-slide tr:nth-child(2n) {
        background-color: white;
    }
    .remark-slide thead, .remark-slide tr:nth-child(n) {
        background-color: white;
    }
```

## Next week `r top_icon("calendar-week")`


### Our final quiz is on April 29th

### Final briefing note is due on May 7th and this is a hard deadline


---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Rent Control

---
### In this chapter (Ch 7)

In this chapter we will examine a few policies that seek to make housing more affordable and accessible. We will look at:

1. Rent control
2. Housing subsidies
3. Minimum dwelling size

---
### Why housing policies?

#### Having housing is crucial, families without it face significant challenges to health (physical and mental), safety, and employment (to name a few)

#### Many families face challenges either getting housing at all, or getting adequate housing

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture1.png"))
```

---
## Can rent control help?

### Rent control creates legal restrictions to what can be charged in rent

### Rent control creates a price ceiling – price cannot increase above this ceiling

### The goal is to make housing rent cheaper by forcing rents lower

---
## Types of rent control laws

####  Rent control laws differ by region but they typically limit the rate of increase of rent for a dwelling during a tenant’s period of residence

####  So if you rent a property that has rent control, your landlord can only increase rent up to the maximum percent increase per year

####  But if you move out and the landlord rents to someone else, she can then increase the rent before she puts your old dwelling back on the market (called “vacancy decontrol”)

####  New buildings are often exempted from rent control laws

---
## Types of rent control laws

### While most rent control laws have this feature (“vacancy decontrol”), some don’t

### Laws that do not allow vacancy decontrol (landlords can’t re-adjust rents when tenants leave) are deemed to be strong rent control

### Strong rent control laws: CA (West Hollywood, Santa Monica, Berkeley, East Palo Alto, Cotati) 

### But these strong rent control laws are unenforceable since 1996 after the Costa-Hawkins Rental Housing Act

---
## What does economics say about Rent Control?

#### Two types of economics research will help us determine the effects of rent control:

1. Predictions from economic models
2. Empirical evidence (e.g., difference-in-difference statistical studies)

We will look at both sets of evidence, starting with (1)

---
## Predictions from economic models

### There are different models of the economy that will generate different predictions of the impact of rent control.

### Consider two extremes:
1. Percent competition
2. Monopoly

### Under perfect competition, rent control will have adverse consequences, making it undesirable

### Under a monopoly, there is a greater likelihood for beneficial outcomes

---
## Rent control under perfect competition

#### Under perfect competition, the market is extremely competitive

#### No firms have market power

#### No consumers have market power

#### Market price is determined by supply and demand and is exogenous

---
## Rent control: Long-run vs. short-run

.pull-left[
#### Housing developers can’t build new housing immediately, so housing supply in the short run is fixed (vertical line)

#### Equilibrium occurs where the downward sloping demand curve intersects the short run supply curve
]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\begin{tikzpicture}[scale=0.8]
  % Axis
  \draw[->] (0,0) -- (5,0) node[right] {$Q$};
  \draw[->] (0,0) -- (0,5) node[above] {$P$};

  % Demand Curve
  \draw[thick] (0.5,4.5) -- (4.5,0.5) node[right] {Demand};

  % Supply Curve (Short Run)
  \draw[thick] (2.5,0.5) -- (2.5,4.5) node[above right] {Supply (short run)};

  % Equilibrium Point
  \filldraw (2.5,2.5) circle (2pt);
  \draw[dashed] (2.5,0) -- (2.5,2.5) node[midway, below left] {$Q^*_{\text{short}}$};
  \draw[dashed] (0,2.5) -- (2.5,2.5) node[midway, above left] {$P^*_{\text{short}}$};

\end{tikzpicture}
```
]

---
## Rent control: Long-run vs. short-run

.pull-left[
In the long run, housing developers can change the housing stock by building more units (S1 to S2) or by letting some units fall into disrepair (S1 to S0).

Since populations are increasing in more cities, it’s usually a matter of how much housing gets built in the long run (a lot vs a little)
]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\begin{tikzpicture}[scale=1.2]

  % Axes
  \draw[->] (0,0) -- (6,0) node[right] {$Q$};
  \draw[->] (0,0) -- (0,5) node[above] {$P$};

  % Vertical Supply Curves
  \draw[thick] (2,0.5) -- (2,4.5) node[above] {$S_0$};
  \draw[thick] (3,0.5) -- (3,4.5) node[above] {$S_1$};
  \draw[thick] (4,0.5) -- (4,4.5) node[above] {$S_2$};

  % Arrows showing movement
  \draw[->, thick] (2.8,2.5) -- (2.2,2.5);
  \draw[->, thick] (3.2,2.5) -- (3.8,2.5);

\end{tikzpicture}
```
]

---
## Rent control: Long-run vs. short-run

.pull-left[
#### The long run supply curve is upward sloping, as in the long run more housing gets built when prices are higher
]
.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\begin{tikzpicture}[scale=1.2]

  % Axes
  \draw[->] (0,0) -- (6,0) node[right] {$Q$};
  \draw[->] (0,0) -- (0,5) node[above] {$P$};

  % Demand and Supply Curves
  \draw[thick] (1,4.5) -- (5,0.5) node[below right] {Demand};
  \draw[thick] (1,0.5) -- (5,4.5) node[above right] {Supply (long run)};

  % Equilibrium point
  \filldraw (3,2.5) circle (2pt);

  % Dashed lines for P* and Q*
  \draw[dashed] (0,2.5) -- (3,2.5) node[midway, above left] {$P^*_{\text{long}}$};
  \draw[dashed] (3,0) -- (3,2.5) node[midway, below right] {$Q^*_{\text{long}}$};

\end{tikzpicture}

```
]

---
## Addition of rent control

#### Rent control operates as a price ceiling: price cannot go above the ceiling

#### This price ceiling can either be binding or non-binding

#### Binding: The price ceiling is less than the equilibrium rent

#### When binding, rent control has an effect on the market

#### Non-Binding: The price ceiling is equal to or greater than the equilibrium rent

#### No effect on the market (for now)

---
## Short run effects of rent control


.pull-left[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{decorations.pathreplacing}

\begin{tikzpicture}[scale=1.2]
    % Axes
    \draw[->] (0,0) -- (0,6) node[left] {$P$};
    \draw[->] (0,0) -- (6,0) node[right] {$Q$};
    
    % Title - larger and bold
    \node[align=center, font=\large\bfseries] at (2.5,5.5) {BINDING};
    
    % Vertical supply line
    \draw[thick] (3,0) -- (3,6) node[above right] {$S^{short}_{run}$};
    
    % Demand curve
    \draw[thick] (1,5) -- (5,1) node[right] {$D$};
    
    % Price control line - P_RC
    \draw[dashed, blue] (0,2.2) -- (3.8,2.2);
    \node[left, blue] at (0,2.2) {$P_{RC}$};
    
    % Q labels at the bottom
    \node[below] at (3,0) {$Q_0$};
    
    % Find the Q_D point where P_RC intersects demand curve (D)
    % Since D goes from (1,5) to (5,1), we can calculate the intersection with P_RC at y=2.2
    % Using the line equation: y = 5 - (4/4)(x-1) = 5 - (x-1) = 6-x
    % When y = 2.2, we get 2.2 = 6-x → x = 3.8
    
    % Quantity demanded at P_RC (which exceeds supply)
    \draw[dotted] (3.8,2.2) -- (3.8,0);
    \node[below] at (3.8,0) {$Q_D$};
    
    % Shortage notation with curly brace
    \draw [decorate, decoration={brace, amplitude=5pt, mirror}] (3,0) -- (3.8,0);
    \node[below] at (3.4,-0.3) {shortage};
\end{tikzpicture}
```
]


.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{decorations.pathreplacing}

\begin{tikzpicture}[scale=1.2]
    % Axes
    \draw[->] (0,0) -- (0,6) node[left] {$P$};
    \draw[->] (0,0) -- (6,0) node[right] {$Q$};
    
    % Title - larger and bold
    \node[align=center, font=\large\bfseries] at (2.5,5.5) {NON-BINDING};
    
    % Vertical supply line
    \draw[thick] (3,0) -- (3,6) node[above right] {$S^{short}_{run}$};
    
    % Demand curve
    \draw[thick] (1,5) -- (5,1) node[right] {$D$};
    
    % Price control line (ABOVE equilibrium - non-binding case)
    \draw[dashed, blue] (0,4.5) -- (3,4.5);
    \node[left, blue] at (0,4.5) {$P_{RC}$};
\end{tikzpicture}

```
]

---
## In the long run

.pull-left[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{decorations.pathreplacing}

\begin{tikzpicture}[scale=1.2]
    % Axes
    \draw[->] (0,0) -- (0,6) node[left] {$P$};
    \draw[->] (0,0) -- (6,0) node[right] {$Q$};
    
    % Title - larger and bold
    \node[align=center, font=\large\bfseries] at (2.5,5.5) {BINDING};
    
    % Normal upward-sloping supply line
    \draw[thick] (1,1) -- (5,5) node[right] {$S$};
    
    % Demand curve
    \draw[thick] (1,5) -- (5,1) node[right] {$D$};
    
    % Equilibrium point (intersection of S and D)
    % Solving for the intersection:
    % S: y = 1 + (4/4)(x-1) = 1 + (x-1) = x
    % D: y = 5 - (x-1) = 6 - x
    % Setting equal: x = 6 - x
    % 2x = 6
    % x = 3
    % y = 3
    \filldraw[black] (3,3) circle (1.5pt);
    
    % Price control line - P_RC (below equilibrium)
    \draw[dashed, blue] (0,2.2) -- (3.8,2.2);
    \node[left, blue] at (0,2.2) {$P_{RC}$};
    
    % Quantity supplied at P_RC
    % When P = 2.2, we solve for Q using supply equation
    % 2.2 = x
    % x = 2.2
    \draw[dotted] (2.2,2.2) -- (2.2,0);
    \node[below] at (2.2,0) {$Q_S$};
    
    % Quantity demanded at P_RC
    % When P = 2.2, we solve for Q using demand equation
    % 2.2 = 6 - x
    % x = 3.8
    \draw[dotted] (3.8,2.2) -- (3.8,0);
    \node[below] at (3.8,0) {$Q_D$};
    
    % Shortage notation with curly brace
    \draw [decorate, decoration={brace, amplitude=5pt, mirror}] (2.2,0) -- (3.8,0);
    \node[below] at (3.0,-0.3) {shortage};
\end{tikzpicture}
```
]


.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{decorations.pathreplacing}

\begin{tikzpicture}[scale=1.2]
    % Axes
    \draw[->] (0,0) -- (0,6) node[left] {$P$};
    \draw[->] (0,0) -- (6,0) node[right] {$Q$};
    
    % Title - larger and bold
    \node[align=center, font=\large\bfseries] at (2.5,5.5) {NON-BINDING};
    
    % Normal upward-sloping supply line
    \draw[thick] (1,1) -- (5,5) node[right] {$S$};
    
    % Demand curve
    \draw[thick] (1,5) -- (5,1) node[right] {$D$};
    
    % Equilibrium point (intersection of S and D)
    \filldraw[black] (3,3) circle (1.5pt);
    
    % Price control line (ABOVE equilibrium - non-binding case)
    \draw[dashed, blue] (0,3.8) -- (3.8,3.8);
    \node[left, blue] at (0,3.8) {$P_{RC}$};
\end{tikzpicture}

```
]

---
## Short run vs long run shortage


```{tikz, fig.ext = 'png', echo = FALSE, fig.align='center', out.width='50%'}
\usetikzlibrary{decorations.pathreplacing}

\begin{tikzpicture}[scale=0.7]
    % Axes
    \draw[->] (0,0) -- (0,6) node[left] {$P$};
    \draw[->] (0,0) -- (6,0) node[right] {$Q$};
        
    % Short-run upward-sloping supply line
    \draw[thick] (1,1) -- (5,5) node[right] {$S_{LR}$};
    
    % Long-run perfectly inelastic supply line (moved to Q = 3)
    \draw[thick, red] (3,0) -- (3,6) node[above] {$S_{SR}$};
    
    % Demand curve
    \draw[thick] (1,5) -- (5,1) node[right] {$D$};
    
    % Equilibrium point (intersection of S and D)
    \filldraw[black] (3,3) circle (1.5pt);
    
    % Price control line - P_RC (below equilibrium)
    \draw[dashed, blue] (0,2.2) -- (3.8,2.2);
    \node[left, blue] at (0,2.2) {$P_{RC}$};
    
    % Short-run quantity supplied at P_RC
    \draw[dotted] (2.2,2.2) -- (2.2,0);
    \node[below=1pt] at (2.2,0) {\scriptsize $Q_{S}^{SR}$};
    
    % Long-run quantity supplied at P_RC (now at Q = 3)
    \draw[dotted, red] (3,2.2) -- (3,0);
    \node[below=1pt] at (3,0) {\scriptsize $Q_{S}^{LR}$};
    
    % Quantity demanded at P_RC
    \draw[dotted] (3.8,2.2) -- (3.8,0);
    \node[below=1pt] at (3.8,0) {\scriptsize $Q_D$};
\end{tikzpicture}
```

---
## Quality of housing

### In addition to a decrease in the quantity of housing in the long run, there is also a decrease in housing quality

### Since landlords get less in rent and face excess demand for their housing, they are less likely to make necessary repairs to buildings

### In competitive markets that are unregulated, price is used as a mechanism to allocate the good to those who are most willing to pay

### With rent control, there is excess demand. Landlords can’t increase rents to clear the market

### How is housing allocated?  

---
## Illegal fees and bribes

### Attempt to bypass rent control by charging fees (usually illegal): e.g., key pickup fee, “deposits” beyond a security deposit

### Prospective tenants may also try to become the favored tenant by offering side payments

---
## Discrimination

### With excess demand, landlords can be picky about who they want to rent to

### Landlords may screen out applicants based on perceived credit, socio-economic status, family type, pets, race, language, religion, sexual orientation, etc

### Those in poverty who need the cheaper housing are the most likely to be screened out

### While discrimination against protected groups is often illegal, it is not well enforced and not all groups are protected (e.g., LGBT)

---
## Inefficient allocation of housing

### With allocation of housing not determined based on price, there is scope for some individuals who have weaker preferences for housing to consume more housing than others who would have stronger preferences

### For example, suppose the equilibrium rent for an apartment would be $1,000 but under rent control it is $900. Suppose Steve is willing to pay $1,000

### With the rent being $900, there are some individuals who would try to rent the apartment who have a willingness to pay of $900 to $999. If they get the apartment over Steve then they get less consumer surplus than Steve, because the apartment is worth more to him

---
## Who benefits from rent control?

### The hope would be that rent control would allow poor families to pay less in rent, thus partially alleviating their poverty

### But those lucky enough to get cheaper housing under rent control aren’t necessarily those who need it most

### Accoring to Gyourko and Linneman (1989), the beneficiaries of rent control are varied and the relationship is non-systematic. Beneficiaries aren’t from any particular social group


---
## Inefficient mobility

### Rent control can also encourage individuals and families to stay in apartments longer than they would normally, because if they move to rent a new place, the new places may have higher rent because landlords can increase the rent once individuals move out

### Thus some moves that would be efficient don’t happen. This reduces social welfare as there is some mis-match between where people live and where they should live

---
## Effects of Rent Control Under Monopoly

There are two models of the economy that represent different extremes:

1. Perfect competition

2. Monopoly

Perfect competition predicts many negative effects of rent control. But under a monopoly, rent control isn’t so bad

---

```{tikz, fig.ext = 'png', echo=FALSE, fig.align='center', out.width='60%'}
\begin{tikzpicture}[scale=1]
    % Axes
    \draw (0,5.8) node [left] {$P$} -- (0,0) node [left] {$0$} -- (5,0) node [right] {$Q$};
    
    % Labels for quantity
    \node [below] at (1.5,0) {$Q_M$};
    \node [below] at (2.4,0) {$Q^*$};

    % Labels for price
    \node [left] at (0,2) {$P^*$};
    \node [left] at (0,3) {$P_M$};

    % MC curve (Long-run supply)
    \draw[thick] (0,1) .. controls (2.3,1.6) and (4,3) .. (4.6,4.5) node [above right] {$MC$};

    % Demand curve
    \draw[thick] (0,5.6) .. controls (1.5,2.75) and (2.4,1.6) .. (4.4,0.4) node [right] {Demand};

    % MR curve
    \draw[thick, dashed] (0,5.6) .. controls (0.8,3) and (1.5,1.5) .. (2.1,0.5) node [below] {MR};

    % Vertical line at Q_M
    \draw[dotted, thick, red] (1.5,0) -- (1.5,3);

    % Vertical line at Q^*
    \draw[dotted, thick, red] (2.4,0) -- (2.4,2);

    % Horizontal lines for price
    \draw[dotted, thick, red] (0,3) -- (1.5,3);
    \draw[dotted, thick, red] (0,2) -- (2.4,2);

    % Points
    \filldraw[red] (1.5,3) circle (1.5pt); % Monopoly equilibrium
    \filldraw[red] (2.4,2) circle (1.5pt); % Competitive equilibrium
\end{tikzpicture}


```

---
## Which case is more realistic?

### In most cities, there are many landlords and many tenants, so the market is very competitive

### Thus the perfect competition model is closer to reality, and the predictions of that model are more plausible

### But there are some cities where companies have lots of market power because they own all or most of the housing (e.g., Irvine Company in Irvine, CA). In this case, rent control might get prices closer to the efficient level if rent control is applied correctly

```{r gen_pdf, include = FALSE, cache = FALSE, eval = TRUE}
library(renderthis)
# https://hhadah.github.io/xaringan/index.html

to_pdf(from = "~/Documents/GiT/urban-econ/slides/11-housing/05-rent-control/rent-control.html", 
       to = "~/Documents/GiT/urban-econ/slides/01-week1/tue/01-class1.pdf")
to_pdf(from = "~/Documents/GiT/urban-econ/slides/11-housing/05-rent-control/rent-control.html")
```
