---
title: "Urban Economics"
subtitle: "Rent Control"
author: "Hussain Hadah (he/him)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: ["default", "assets/tulane-fonts.css", "assets/tulane.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: true # show a title slide with YAML information
    includes:
      in_header: "assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
layout: true
<div style="position: absolute;left:20px;bottom:5px;color:black;font-size: 12px;">`r rmarkdown::metadata$author` (Tulane) | `r rmarkdown::metadata$subtitle` | `r format(Sys.time(), '%d %B %Y')`</div>

<!--- `r rmarkdown::metadata$subtitle` | `r format(Sys.time(), '%d %B %Y')`-->

---
class: title-slide
background-image: url("assets/TulaneLogo-white.svg"), url("assets/title-image1.jpg")
background-position: 10% 90%, 100% 50%
background-size: 160px, 50% 100%
background-color: #0148A4
# .text-shadow[.white[Outline for Today]]

<ol>
    <li><h4 class="white"> </h4></li>
</ol>

---

```{r, load_refs, echo=FALSE, cache=FALSE, message=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'authoryear', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("assets/example.bib", check = FALSE)
top_icon = function(x) {
  icons::icon_style(
    icons::fontawesome(x),
    position = "fixed", top = 10, right = 10
  )
}
```

```{r setup, include=FALSE}
# xaringanExtra::use_scribble() ## Draw on slides. Requires dev version of xaringanExtra.

options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=4, fig.width=6,
  out.width="748px", out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
```


```{r, cache=FALSE, message=FALSE, warning=FALSE, include=TRUE, eval=TRUE, results=FALSE, echo=FALSE, tidy.opts = list(width.cutoff = 50), tidy = TRUE}
## Load and install the packages that we'll be using today
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tictoc, parallel, pbapply, future, future.apply, furrr, RhpcBLASctl, memoise, here, foreign, mfx, tidyverse, hrbrthemes, estimatr, ivreg, fixest, sandwich, lmtest, margins, vtable, broom, modelsummary, stargazer, fastDummies, recipes, dummy, gplots, haven, huxtable, kableExtra, gmodels, survey, gtsummary, data.table, tidyfast, dtplyr, microbenchmark, ggpubr, tibble, viridis, wesanderson, censReg, rstatix, srvyr, formatR, sysfonts, showtextdb, showtext, thematic, sampleSelection, RefManageR, DT, googleVis, png, grid)
# devtools::install_github("thomasp85/patchwork")
remotes::install_github("mitchelloharawild/icons")
remotes::install_github("ROpenSci/bibtex")

# devtools::install_github("ajdamico/lodown")
## My preferred ggplot2 plotting theme (optional)
## https://github.com/hrbrmstr/hrbrthemes
# scale_fill_ipsum()
# scale_color_ipsum()
font_add_google("Fira Sans", "firasans")
font_add_google("Fira Code", "firasans")

showtext_auto()

theme_customs <- theme(
  text = element_text(family = 'firasans', size = 16),
  plot.title.position = 'plot',
  plot.title = element_text(
    face = 'bold', 
    colour = thematic::okabe_ito(8)[6],
    margin = margin(t = 2, r = 0, b = 7, l = 0, unit = "mm")
  ),
)

colors <-  thematic::okabe_ito(5)

# theme_set(theme_minimal() + theme_customs)
theme_set(hrbrthemes::theme_ipsum() + theme_customs)
## Set master directory where all sub-directories are located
mdir <- "/Users/hhadah/Documents/GiT/urban-econ/slides/01-week1/tue"

## Set working directory

# COLOR PALLETES
library(paletteer) 
# paletteer_d("awtools::a_palette")
# paletteer_d("suffrager::CarolMan")

### COLOR BLIND PALLETES
#paletteer_d("colorblindr::OkabeIto")
#paletteer_d("colorblindr::OkabeIto_black")
# paletteer_d("colorBlindness::paletteMartin")
# paletteer_d("colorBlindness::Blue2DarkRed18Steps")
# paletteer_d("colorBlindness::SteppedSequential5Steps")
# paletteer_d("colorBlindness::PairedColor12Steps")
# paletteer_d("colorBlindness::ModifiedSpectralScheme11Steps")
colorBlindness <- paletteer_d("colorBlindness::Blue2Orange12Steps")
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# scale_colour_paletteer_d("colorBlindness::ModifiedSpectralScheme11Steps", dynamic = FALSE)
# To use for fills, add
  scale_fill_manual(values="colorBlindness::Blue2Orange12Steps")

# To use for line and point colors, add
  scale_colour_manual(values="colorBlindness::Blue2Orange12Steps")
  #<a><button>[Click me](#sources)</button></a>
```

```{css, echo=F}
    /* Table width = 100% max-width */

    .remark-slide table{
        width: auto !important; /* Adjusts table width */
    }

    /* Change the background color to white for shaded rows (even rows) */

    .remark-slide thead, .remark-slide tr:nth-child(2n) {
        background-color: white;
    }
    .remark-slide thead, .remark-slide tr:nth-child(n) {
        background-color: white;
    }
```

## Next week `r top_icon("calendar-week")`


### Our final quiz is on April 29th

### Final briefing note is due on May 7th and this is a hard deadline


---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Rent Control

---
### In this chapter (Ch 7)

In this chapter we will examine a few policies that seek to make housing more affordable and accessible. We will look at:

1. Rent control
2. Housing subsidies
3. Minimum dwelling size

---
### Why housing policies?

#### Having housing is crucial, families without it face significant challenges to health (physical and mental), safety, and employment (to name a few)

#### Many families face challenges either getting housing at all, or getting adequate housing

```{r, echo=FALSE, fig.align='center'}
grid.raster(readPNG("assets/picture1.png"))
```

---
## Can rent control help?

### Rent control creates legal restrictions to what can be charged in rent

### Rent control creates a price ceiling – price cannot increase above this ceiling

### The goal is to make housing rent cheaper by forcing rents lower

---
## Types of rent control laws

####  Rent control laws differ by region but they typically limit the rate of increase of rent for a dwelling during a tenant’s period of residence

####  So if you rent a property that has rent control, your landlord can only increase rent up to the maximum percent increase per year

####  But if you move out and the landlord rents to someone else, she can then increase the rent before she puts your old dwelling back on the market (called “vacancy decontrol”)

####  New buildings are often exempted from rent control laws

---
## Types of rent control laws

### While most rent control laws have this feature (“vacancy decontrol”), some don’t

### Laws that do not allow vacancy decontrol (landlords can’t re-adjust rents when tenants leave) are deemed to be strong rent control

### Strong rent control laws: CA (West Hollywood, Santa Monica, Berkeley, East Palo Alto, Cotati) 

### But these strong rent control laws are unenforceable since 1996 after the Costa-Hawkins Rental Housing Act

---
## What does economics say about Rent Control?

#### Two types of economics research will help us determine the effects of rent control:

1. Predictions from economic models
2. Empirical evidence (e.g., difference-in-difference statistical studies)

We will look at both sets of evidence, starting with (1)

---
## Predictions from economic models

### There are different models of the economy that will generate different predictions of the impact of rent control.

### Consider two extremes:
1. Percent competition
2. Monopoly

### Under perfect competition, rent control will have adverse consequences, making it undesirable

### Under a monopoly, there is a greater likelihood for beneficial outcomes

---
## Rent control under perfect competition

#### Under perfect competition, the market is extremely competitive

#### No firms have market power

#### No consumers have market power

#### Market price is determined by supply and demand and is exogenous

---
## Rent control: Long-run vs. short-run

.pull-left[
#### Housing developers can’t build new housing immediately, so housing supply in the short run is fixed (vertical line)

#### Equilibrium occurs where the downward sloping demand curve intersects the short run supply curve
]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\begin{tikzpicture}[scale=0.8]
  % Axis
  \draw[->] (0,0) -- (5,0) node[right] {$Q$};
  \draw[->] (0,0) -- (0,5) node[above] {$P$};

  % Demand Curve
  \draw[thick] (0.5,4.5) -- (4.5,0.5) node[right] {Demand};

  % Supply Curve (Short Run)
  \draw[thick] (2.5,0.5) -- (2.5,4.5) node[above right] {Supply (short run)};

  % Equilibrium Point
  \filldraw (2.5,2.5) circle (2pt);
  \draw[dashed] (2.5,0) -- (2.5,2.5) node[midway, below left] {$Q^*_{\text{short}}$};
  \draw[dashed] (0,2.5) -- (2.5,2.5) node[midway, above left] {$P^*_{\text{short}}$};

\end{tikzpicture}
```
]

---
## Rent control: Long-run vs. short-run

.pull-left[
In the long run, housing developers can change the housing stock by building more units (S1 to S2) or by letting some units fall into disrepair (S1 to S0).

Since populations are increasing in more cities, it’s usually a matter of how much housing gets built in the long run (a lot vs a little)
]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\begin{tikzpicture}[scale=1.2]

  % Axes
  \draw[->] (0,0) -- (6,0) node[right] {$Q$};
  \draw[->] (0,0) -- (0,5) node[above] {$P$};

  % Vertical Supply Curves
  \draw[thick] (2,0.5) -- (2,4.5) node[above] {$S_0$};
  \draw[thick] (3,0.5) -- (3,4.5) node[above] {$S_1$};
  \draw[thick] (4,0.5) -- (4,4.5) node[above] {$S_2$};

  % Arrows showing movement
  \draw[->, thick] (2.8,2.5) -- (2.2,2.5);
  \draw[->, thick] (3.2,2.5) -- (3.8,2.5);

\end{tikzpicture}
```
]

---
## Rent control: Long-run vs. short-run

.pull-left[
#### The long run supply curve is upward sloping, as in the long run more housing gets built when prices are higher
]
.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\begin{tikzpicture}[scale=1.2]

  % Axes
  \draw[->] (0,0) -- (6,0) node[right] {$Q$};
  \draw[->] (0,0) -- (0,5) node[above] {$P$};

  % Demand and Supply Curves
  \draw[thick] (1,4.5) -- (5,0.5) node[below right] {Demand};
  \draw[thick] (1,0.5) -- (5,4.5) node[above right] {Supply (long run)};

  % Equilibrium point
  \filldraw (3,2.5) circle (2pt);

  % Dashed lines for P* and Q*
  \draw[dashed] (0,2.5) -- (3,2.5) node[midway, above left] {$P^*_{\text{long}}$};
  \draw[dashed] (3,0) -- (3,2.5) node[midway, below right] {$Q^*_{\text{long}}$};

\end{tikzpicture}

```
]

---
## Addition of rent control

#### Rent control operates as a price ceiling: price cannot go above the ceiling

#### This price ceiling can either be binding or non-binding

#### Binding: The price ceiling is less than the equilibrium rent

#### When binding, rent control has an effect on the market

#### Non-Binding: The price ceiling is equal to or greater than the equilibrium rent

#### No effect on the market (for now)

---
title: "Rent Control Analysis"
output: 
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Short run effects of rent control

.pull-left[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{decorations.pathreplacing}

\begin{tikzpicture}[scale=1.2]
    % Axes
    \draw[->] (0,0) -- (0,6) node[left] {$P$};
    \draw[->] (0,0) -- (6,0) node[right] {$Q$};
    
    % Title - larger and bold
    \node[align=center, font=\large\bfseries] at (2.5,5.5) {BINDING};
    
    % Vertical supply line
    \draw[thick] (3,0) -- (3,6) node[above right] {$S^{short}_{run}$};
    
    % Demand curve
    \draw[thick] (1,5) -- (5,1) node[right] {$D$};
    
    % Equilibrium price line (higher than control)
    \draw[dashed] (0,3.7) -- (3,3.7);
    \node[left] at (0,3.7) {$P_e$};
    
    % Price control line (BELOW equilibrium - binding case) - blue color
    \draw[dashed, blue] (0,2.2) -- (3,2.2);
    \node[left, blue] at (0,2.2) {$P_c$};
    
    % Q labels at the bottom
    \node[below] at (3,0) {$Q_0$};
    
    % Quantity demanded at Pc (which exceeds supply)
    \draw[dotted] (4.2,2.2) -- (4.2,0);
    \node[below] at (4.2,0) {$Q_D$};
    
    % Shortage notation with curly brace
    \draw [decorate, decoration={brace, amplitude=5pt, mirror}] (3,0) -- (4.2,0);
    \node[below] at (3.6,-0.3) {shortage};
\end{tikzpicture}
```
]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{decorations.pathreplacing}

\begin{tikzpicture}[scale=1.2]
    % Axes
    \draw[->] (0,0) -- (0,6) node[left] {$P$};
    \draw[->] (0,0) -- (6,0) node[right] {$Q$};
    
    % Title - larger and bold
    \node[align=center, font=\large\bfseries] at (2.5,5.5) {NOT BINDING};
    
    % Vertical supply line
    \draw[thick] (3,0) -- (3,6) node[above right] {$S^{short}_{run}$};
    
    % Demand curve
    \draw[thick] (1,5) -- (5,1) node[right] {$D$};
    
    % Price control line (ABOVE equilibrium - blue color)
    \draw[dashed, blue] (0,3.7) -- (3,3.7);
    \node[left, blue] at (0,3.7) {$P_c$};
    
    % Equilibrium price line (lower than control)
    \draw[dashed] (0,2.2) -- (3,2.2);
    \node[left] at (0,2.2) {$P_e$};
    
    % Q labels at the bottom
    \node[below] at (3,0) {$Q_e$};
    
    % Market clears notation
    \node[below] at (3.6,-0.3) {market clears};
\end{tikzpicture}
```
]


```{r gen_pdf, include = FALSE, cache = FALSE, eval = FALSE}
library(renderthis)
# https://hhadah.github.io/xaringan/index.html

to_pdf(from = "~/Documents/GiT/urban-econ/slides/01-week1/tue/01-class1.html", 
       to = "~/Documents/GiT/urban-econ/slides/01-week1/tue/01-class1.pdf")
to_pdf(from = "~/Documents/GiT/urban-econ/slides/01-week1/tue/01-class1.html")
```
