---
title: "Urban Economics"
subtitle: "Introduction to Class"
author: "Hussain Hadah (he/him)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: ["default", "assets/tulane-fonts.css", "assets/tulane.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: true # show a title slide with YAML information
    includes:
      in_header: "assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
layout: true
<div style="position: absolute;left:20px;bottom:5px;color:black;font-size: 12px;">`r rmarkdown::metadata$author` (Tulane) | `r rmarkdown::metadata$subtitle` | `r format(Sys.time(), '%d %B %Y')`</div>

<!--- `r rmarkdown::metadata$subtitle` | `r format(Sys.time(), '%d %B %Y')`-->

---

```{r, load_refs, echo=FALSE, cache=FALSE, message=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'authoryear', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("assets/example.bib", check = FALSE)
top_icon = function(x) {
  icons::icon_style(
    icons::fontawesome(x),
    position = "fixed", top = 10, right = 10
  )
}
```

```{r setup, include=FALSE}
# xaringanExtra::use_scribble() ## Draw on slides. Requires dev version of xaringanExtra.

options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=4, fig.width=6,
  out.width="748px", out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
```


```{r, cache=FALSE, message=FALSE, warning=FALSE, include=TRUE, eval=TRUE, results=FALSE, echo=FALSE, tidy.opts = list(width.cutoff = 50), tidy = TRUE}
## Load and install the packages that we'll be using today
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tictoc, parallel, pbapply, future, future.apply, furrr, RhpcBLASctl, memoise, here, foreign, mfx, tidyverse, hrbrthemes, estimatr, ivreg, fixest, sandwich, lmtest, margins, vtable, broom, modelsummary, stargazer, fastDummies, recipes, dummy, gplots, haven, huxtable, kableExtra, gmodels, survey, gtsummary, data.table, tidyfast, dtplyr, microbenchmark, ggpubr, tibble, viridis, wesanderson, censReg, rstatix, srvyr, formatR, sysfonts, showtextdb, showtext, thematic, sampleSelection, RefManageR, DT, googleVis, png, grid)
# devtools::install_github("thomasp85/patchwork")
remotes::install_github("mitchelloharawild/icons")
remotes::install_github("ROpenSci/bibtex")

# devtools::install_github("ajdamico/lodown")
## My preferred ggplot2 plotting theme (optional)
## https://github.com/hrbrmstr/hrbrthemes
# scale_fill_ipsum()
# scale_color_ipsum()
font_add_google("Fira Sans", "firasans")
font_add_google("Fira Code", "firasans")

showtext_auto()

theme_customs <- theme(
  text = element_text(family = 'firasans', size = 16),
  plot.title.position = 'plot',
  plot.title = element_text(
    face = 'bold', 
    colour = thematic::okabe_ito(8)[6],
    margin = margin(t = 2, r = 0, b = 7, l = 0, unit = "mm")
  ),
)

colors <-  thematic::okabe_ito(5)

# theme_set(theme_minimal() + theme_customs)
theme_set(hrbrthemes::theme_ipsum() + theme_customs)
## Set master directory where all sub-directories are located
mdir <- "/Users/hhadah/Documents/GiT/urban-econ/slides/01-week1/tue"

## Set working directory

# COLOR PALLETES
library(paletteer) 
# paletteer_d("awtools::a_palette")
# paletteer_d("suffrager::CarolMan")

### COLOR BLIND PALLETES
#paletteer_d("colorblindr::OkabeIto")
#paletteer_d("colorblindr::OkabeIto_black")
# paletteer_d("colorBlindness::paletteMartin")
# paletteer_d("colorBlindness::Blue2DarkRed18Steps")
# paletteer_d("colorBlindness::SteppedSequential5Steps")
# paletteer_d("colorBlindness::PairedColor12Steps")
# paletteer_d("colorBlindness::ModifiedSpectralScheme11Steps")
colorBlindness <- paletteer_d("colorBlindness::Blue2Orange12Steps")
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# scale_colour_paletteer_d("colorBlindness::ModifiedSpectralScheme11Steps", dynamic = FALSE)
# To use for fills, add
  scale_fill_manual(values="colorBlindness::Blue2Orange12Steps")

# To use for line and point colors, add
  scale_colour_manual(values="colorBlindness::Blue2Orange12Steps")
  #<a><button>[Click me](#sources)</button></a>
```

```{css, echo=F}
    /* Table width = 100% max-width */

    .remark-slide table{
        width: auto !important; /* Adjusts table width */
    }

    /* Change the background color to white for shaded rows (even rows) */

    .remark-slide thead, .remark-slide tr:nth-child(2n) {
        background-color: white;
    }
    .remark-slide thead, .remark-slide tr:nth-child(n) {
        background-color: white;
    }
```

## About Me `r top_icon("user")`

- My name is [Hussain Hadah](https://hussainhadah.com)  

- Email: [hhadah@tulane.edu](mailto:hhadah@tulane.edu)

- I am an labor economist

  - I study topics on discrimination, racial and ethnic identities, mental health, and immigration

- I love to run

  - You can find me running in Audubon Park with my dog Bill

```{r example1, echo=FALSE, warning=FALSE, out.width="30%", fig.show='hold', fig.align='right'}
# Read the image file
img <- readPNG("assets/bill.png")

# Plot the image
grid.raster(img)
```

- Student hours/office hours are by appointment using  [https://hussainhadah.youcanbook.me/](https://hussainhadah.youcanbook.me/)

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Overview `r top_icon("list-ul")`

---
## This Week `r top_icon("calendar-week")`

1. Introduction to the course

2. Geographical data

### Readings `r icons::fontawesome("book-open")`

- Syllabus

- OSullivan - Census Definitions

### What's due? `r icons::fontawesome("exclamation")`

- Scavenger Hunt---no due date and you can redo to improve your grade

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Introducing Urban Economics `r top_icon("list-ul")`

---

## What is Urban Economics?

.blockquote[
### `r icons::fontawesome("comment-dots")` .brand-crawfest[Urban economics]

- .black[Introduces space into economics models and studies the location of economic activity]
]

--
- A lot of fields in economics overlap with .brand-crawfest[urban economics]

  - For example, economics of crime overlaps with labor economics and .brand-crawfest[urban economics]


---

## What is Urban Economics?

- Given how broad .brand-crawfest[urban economics] is, I will focus on the following topics

  - Geographical data
  - Agglomeration economies and the economics of cities
  - Economic development and tax incentives
  - Crime
  - Gender-based violence
  - Housing and housing policy
    - Zoning laws (maybe?)
  - Neighborhood effects
  - Transportation

---

## What is Urban Economics?

- This class will be a mix of .brand-medallion[theory] and .brand-blue[empirical] research on urban economics

  - .brand-medallion[Theory]: Applications of mathematical models

  - .brand-blue[Empirical]: Applications of statistical analysis

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Going Over the Syllabus `r top_icon("list-ul")`

---
## Prerequisites

- This class assumes that you have taken: 

  1. Introductory microeconomics (ECON 1010)
  2. High school algebra (solve for X, graph a line, etc.)
  3. High school statistics (calculating medi- ans, ideally remembering the basics of hypothesis testing, etc.)

- It is helpful if you are familiar with statistics and econometrics, but it is not required
  - Knowing what a standard error and confidence interval are is helpful
  - The basics of linear regression

---
## Readings

The required book is Brueckner, Jan. 2011. Lectures on Urban Economics. Cambridge: MIT University Press

- Half of the readings are from the textbook

- The other half are from other sources or my lectures

- Some of the readings are technical pieces from economics journals

- The degree to which you to be familiar with the details of a paper will be clear from the emphasis given to the paper in lecture or will be clear based on instructions I give you

--
- I will either send emails about the readings or announce in class

- It is your responsibility to keep up with the announcements, not doing so will affect your grade

  - Many in-class activities may require that you have read the readings before class

---
## Grading

.left-column[
- Quizzes(Best $3$ out of $4 \times 20\%=60\%$)
- Group Briefing Notes (Best $2$ out of $3 \times 10\%=20\%$)
- Other Activities ($20%$) (Three activities dropped)
]

.right-column[
| Grade | Percentage Range |
|:-----:|:----------------:|
| A     | 93% to 100%      |
| A-    | 90% to 92.99%    |
| B+    | 87% to 89.99%    |
| B     | 83% to 86.99%    |
| B-    | 80% to 82.99%    |
| C+    | 77% to 79.99%    |
| C     | 73% to 76.99%    |
| C-    | 70% to 72.99%    |
| D+    | 67% to 69.99%    |
| D     | 63% to 66.99%    |
| D-    | 60% to 62.99%    |
| F     | 0% to 59.99%     |
]

---
## Quizzes

.pull-left[

- There will be four quizzes

- During class time

- Tentative schedule is Feb. 8, Mar. 5, Mar. 28, and Apr. 25

- Quiz are open book, you can bring notes, documents, computer, etc.

- You cannot talk or work with other students

- You can take the quiz remotely as well


]

--

.pull-right[

- There will be two or three short answer questions

- Zero to two multiple choice questions

]

---
## Group Briefing Notes

.pull-left[

- There will be three group briefing notes

- The goal is to practice:
    - Concise writing
    - Citing sources, and creating a references section in APA format or Chicago Author-Year format
    - Doing a literature review using Google Scholar
    - Reading and summarizing published economics journal articles
    - Coming to reasonable conclusions based on your assessment of the research

]

.pull-right[

- You will form groups of 3-5 students beforehand, or I will create groups for you randomly though a jigsaw activity

- The group will work during class time to write a one-page briefing note, single-spaced, in a Google Doc or OneDrive

- Guidance for the content will be given through handouts that are on Canvas beforehand

- The note will quickly introduce the reader to the topic and summarize the main findings of the research
]

---
## Group Briefing Notes Topics 

.pull-left[
1. To what extent do tax incentives encourage economic development? 
2. Is there racial bias in policing and in the criminal justice system?
3. What are the impacts of rent control on housing markets?

- You will work on first draft in-class
- You can finish on your own time if you did not finish
- There are .red[NO] deadlines to submit the notes---other than the firm end-of-course deadlines
- After Apr. 30, I will grade the notes as your final version


]

.pull-right[
- I will then grade your first draft as if it is a final document, so your grade will be lower

- First round is only temporary

- If you submit the draft before the next class (deadline), you will have chance to make revisions

]

---
## Other Activities

- Lots of research shows that students learn better when they are actively engaged in the learning process

- I will use a variety of activities to engage you in the learning process

- Weight of activity will depend on complexity

- Lowest three will be dropped

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# ADA, Accessibility, Title IX, and Inclusivity Statements and Supports

---
## Making this Class an Inclusive Environment

- I am committed to making this class an inclusive environment for all students

- If you require accommodations, please do the following:

  - Contact me immediately

  - Contact the Goldman Center for Student Accessibility

    - goldman@tulane.edu
    - (504) 862-8433 
    - https://accessibility.tulane.edu/

- I **.red[DO NOT]** need to know the details of your accommodations

- I will **.red[NOT]** ask you for a proof of your accommodations

- Register with the Goldman Center for Student Accessibility as soon as possible

- More details on the syllabus

---
## Children in Class

- I believe that more women and parents should be in academia

- I am happy to have children in class

- More details on the syllabus

- I ask of you to create a welcoming environment for parents and children

---
## Preferred Pronouns

- My preferred pronouns are he/him/his

- Thanks to those that already have it on their accounts

- I recommend everyone to add their preferred pronouns to their accounts

---
## Title IX

- Respect for all people is a core value of Tulane University

- Tulane is committed to providing an environment free of all forms of discrimination and sexual harassment, including sexual assault, domestic and dating violence, and stalking

- If you (or someone you know) has experienced or is experiencing these types of behaviors, know that you are not alone:

  - Resources and support are available: you can learn more at https://allin.tulane.edu/

- Please know that if you choose to confide in me I am required by the university to share your disclosure with Care Connection to the Office of Case Management and Victim Support Services

- You can also make a disclosure yourself, including an anonymous report
  
  - The form is available at tulane.edu/concerns
---
## Lauren’s Promise

- **I will listen and believe you if someone is threatening or harassing you**

- Lauren McCluskey, a 21-year-old honors student athlete, was murdered on Oct. 22, 2018, by a man she briefly dated on the University of Utah campus
- Lauren’s mother, Jill McCluskey, has made it her mission to ensure that no other parent experiences the loss of a child to violence
- We must all do our part to keep our community safe so that this would never happen again

- Learn more at https://laurenmccluskey.org/laurens-promise/

---
## Title IX

```{r title-ix, echo=FALSE, warning=FALSE, out.width="80%", fig.show='hold', fig.align='center'}
# Read the image file
img <- readPNG("/Users/hhadah/Documents/GiT/urban-econ/syllabus/TulaneTitleIX.jpg")

# Plot the image
grid.raster(img)
```

---
class: tulane-blue
background-image: url(assets/TulaneLogo-white.svg)
background-size: 260px
background-position: 5% 95%

# Any questions?

.pull-right[.pull-down[

<a href="mailto:hhadah@tulane.edu">
.white[`r icons::fontawesome("paper-plane")` hhadah@tulane.edu]
</a>

<a href="https://hussainhadah.com/">
.white[`r icons::fontawesome("link")` https://hussainhadah.com/]
</a>

<a href="https://github.com/hhadah">
.white[`r icons::fontawesome("github")` @hhadah]
</a>

<br><br><br>

]]


```{r gen_pdf, include = FALSE, cache = FALSE, eval = TRUE}
library(renderthis)
# https://hhadah.github.io/xaringan/index.html

to_pdf(from = "~/Documents/GiT/urban-econ/slides/01-week1/tue/01-class1.html", 
       to = "~/Documents/GiT/urban-econ/slides/01-week1/tue/01-class1.pdf")
to_pdf(from = "~/Documents/GiT/urban-econ/slides/01-week1/tue/01-class1.html")
```
