---
title: "Explanation of Common Terminology in Causal Studies in (Urban) Economics"
subtitle: "Common Terminology in Causal Studies"
author: "Hussain Hadah (he/him)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: ["default", "assets/tulane-fonts.css", "assets/tulane.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: true # show a title slide with YAML information
    includes:
      in_header: "assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---
layout: true
<div style="position: absolute;left:20px;bottom:5px;color:black;font-size: 12px;">`r rmarkdown::metadata$author` (Tulane) | `r rmarkdown::metadata$subtitle` | `r format(Sys.time(), '%d %B %Y')`</div>

<!--- `r rmarkdown::metadata$subtitle` | `r format(Sys.time(), '%d %B %Y')`-->

---
class: title-slide
background-image: url("assets/TulaneLogo-white.svg"), url("assets/title-image1.jpg")
background-position: 10% 90%, 100% 50%
background-size: 160px, 50% 100%
background-color: #0148A4

```{r, load_refs, echo=FALSE, cache=FALSE, message=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'authoryear', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("assets/example.bib", check = FALSE)
top_icon = function(x) {
  icons::icon_style(
    icons::fontawesome(x),
    position = "fixed", top = 10, right = 10
  )
}
```

```{r setup, include=FALSE}
# xaringanExtra::use_scribble() ## Draw on slides. Requires dev version of xaringanExtra.

options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center",  
  fig.height=4, fig.width=6,
  out.width="748px", out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
```


```{r, cache=FALSE, message=FALSE, warning=FALSE, include=TRUE, eval=TRUE, results=FALSE, echo=FALSE, tidy.opts = list(width.cutoff = 50), tidy = TRUE}
## Load and install the packages that we'll be using today
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tictoc, parallel, pbapply, future, future.apply, furrr, RhpcBLASctl, memoise, here, foreign, mfx, tidyverse, hrbrthemes, estimatr, ivreg, fixest, sandwich, lmtest, margins, vtable, broom, modelsummary, stargazer, fastDummies, recipes, dummy, gplots, haven, huxtable, kableExtra, gmodels, survey, gtsummary, data.table, tidyfast, dtplyr, microbenchmark, ggpubr, tibble, viridis, wesanderson, censReg, rstatix, srvyr, formatR, sysfonts, showtextdb, showtext, thematic, sampleSelection, RefManageR, DT, googleVis, png, grid)
# devtools::install_github("thomasp85/patchwork")
remotes::install_github("mitchelloharawild/icons")
remotes::install_github("ROpenSci/bibtex")

# devtools::install_github("ajdamico/lodown")
## My preferred ggplot2 plotting theme (optional)
## https://github.com/hrbrmstr/hrbrthemes
# scale_fill_ipsum()
# scale_color_ipsum()
font_add_google("Fira Sans", "firasans")
font_add_google("Fira Code", "firasans")

showtext_auto()

theme_customs <- theme(
  text = element_text(family = 'firasans', size = 16),
  plot.title.position = 'plot',
  plot.title = element_text(
    face = 'bold', 
    colour = thematic::okabe_ito(8)[6],
    margin = margin(t = 2, r = 0, b = 7, l = 0, unit = "mm")
  ),
)

colors <-  thematic::okabe_ito(5)

# theme_set(theme_minimal() + theme_customs)
theme_set(hrbrthemes::theme_ipsum() + theme_customs)
## Set master directory where all sub-directories are located
mdir <- "/Users/hhadah/Documents/GiT/urban-econ/slides/06-week6/01-tue"

## Set working directory

# COLOR PALLETES
library(paletteer) 
# paletteer_d("awtools::a_palette")
# paletteer_d("suffrager::CarolMan")

### COLOR BLIND PALLETES
#paletteer_d("colorblindr::OkabeIto")
#paletteer_d("colorblindr::OkabeIto_black")
# paletteer_d("colorBlindness::paletteMartin")
# paletteer_d("colorBlindness::Blue2DarkRed18Steps")
# paletteer_d("colorBlindness::SteppedSequential5Steps")
# paletteer_d("colorBlindness::PairedColor12Steps")
# paletteer_d("colorBlindness::ModifiedSpectralScheme11Steps")
colorBlindness <- paletteer_d("colorBlindness::Blue2Orange12Steps")
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# scale_colour_paletteer_d("colorBlindness::ModifiedSpectralScheme11Steps", dynamic = FALSE)
# To use for fills, add
  scale_fill_manual(values="colorBlindness::Blue2Orange12Steps")

# To use for line and point colors, add
  scale_colour_manual(values="colorBlindness::Blue2Orange12Steps")
  #<a><button>[Click me](#sources)</button></a>
```

```{css, echo=F}
    /* Table width = 100% max-width */

    .remark-slide table{
        width: auto !important; /* Adjusts table width */
    }

    /* Change the background color to white for shaded rows (even rows) */

    .remark-slide thead, .remark-slide tr:nth-child(2n) {
        background-color: white;
    }
    .remark-slide thead, .remark-slide tr:nth-child(n) {
        background-color: white;
    }
```
# .text-shadow[.white[Outline for Today]]

<ol>
    <li><h4 class="white">Introduce Endogeneity <br> ("Chicken or the Egg") Issue in Crime Econ</h4></li>
    <li><h4 class="white">Exogeneity vs. Endogeneity</h4></li>
    <li><h4 class="white">Bias in Causal Inference</h4></li>
</ol>

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Issues in Causal Inference: Endogeneity

---
## Endogeneity in Crime Econ `r top_icon("question")`

.pull-left[
- Suppose you were to compare areas/cities with more police officers to areas/cities with fewer officers to see how crime differs. 
- Do the areas/cities with more officers have less crime?
- But the number of police officers is endogenous to crime.
- Endogenous since crime affects the number of police officers, but police officers affect crime.
- E.g., police officers allocated to high crime areas.
- Increases in crime prompt the hiring of additional officers.
]

.pull-right[
```{r, echo=FALSE, warning=FALSE, out.width="100%", fig.show='hold', fig.align='center', caption="The Rational Criminal by DALL-E"}
# Read the image file
img1 <- readPNG("assets/Picture1.png")

# Plot the image
grid.raster(img1)
```

]

---
## Endogeneity in Crime Econ `r top_icon("question")`

.pull-left[
- Suppose you were to do this comparison anyways…

- Suppose you were to compare areas/cities with more police officers to areas/cities with fewer officers to see how crime differs. 
- Do the areas/cities with more officers have less crime?
- Do you think that by doing this comparison you would overestimate or underestimate the effect of police on reducing crime?
]

.pull-right[
```{r, echo=FALSE, warning=FALSE, out.width="100%", fig.show='hold', fig.align='center', caption="The Rational Criminal by DALL-E"}
# Read the image file
img1 <- readPNG("assets/Picture1.png")

# Plot the image
grid.raster(img1)
```

]

---
## Endogeneity in Crime Econ `r top_icon("question")`

.pull-left[
- Suppose you were to compare areas/cities with more police officers to areas/cities with fewer officers to see how crime differs.
- This would probably underestimate the effect of police on crime, perhaps showing incorrectly that they increase crime, or that their effect on crime reduction is smaller than it actually is.
- Estimates would be negatively biased.
- Why?
]

.pull-right[
```{r, echo=FALSE, warning=FALSE, out.width="100%", fig.show='hold', fig.align='center', caption="The Rational Criminal by DALL-E"}
# Read the image file
img1 <- readPNG("assets/Picture1.png")

# Plot the image
grid.raster(img1)
```

]

---
## Endogeneity in Crime Econ `r top_icon("question")`

.pull-left[
- Estimates would be negatively biased.
- Why?
- Since police are allocated to places with higher crime rates, or more police are deployed when crime increases, there is going to be a positive correlation between the two.
- Mistaking that for a causal relationship will bias the estimate.
- We have to break this endogeneity loop!
]

.pull-right[
```{r, echo=FALSE, warning=FALSE, out.width="100%", fig.show='hold', fig.align='center', caption="The Rational Criminal by DALL-E"}
# Read the image file
img1 <- readPNG("assets/Picture1.png")

# Plot the image
grid.raster(img1)
```

]

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# How to Break the Endogeneity Loop?

---
## Solving the Egg and Chicken Problem: DALL-E

```{r, echo=FALSE, warning=FALSE, out.width="70%", fig.show='hold', fig.align='center', caption="The Rational Criminal by DALL-E"}
# Read the image file
img2 <- readPNG("assets/egg-chicken.png")

# Plot the image
grid.raster(img2)
```

---
## Randomization

- We covered how Differece-in-Difference (DiD) can help us break the endogeneity loop.
- The ideal way to investigate the effect of police on crime would be to randomly allocate areas/cities with more/fewer police officers.
- Likely not possible…
- Is there a way that police are allocated that is “quasi-random”?
- Is there a natural experiment?
- Or, phrased another way, is there a way that police were allocated that was independent from the crime level (i.e. Is there a case where police officers were not allocated based on crime levels?)

---
## Empirical Studies on How Police Affect Crime

- In this course we will cover some neat empirical research articles that investigate how police affect crime using different experimental or “quasi-experimental” methods.
- These are the readings for the Jigsaw activity
- **Levitt, Steven D.** 1997. “Using Electoral Cycles in Police Hiring to Estimate the Effect of Police on
Crime.” American Economic Review, 87(3): 270–290.
- **Sullivan, Christopher M, and Zachary P. O’Keeffe.** 2017. “Evidence that curtailing proactive
policing can reduce major crime.” Nature Human Behaviour, 1(10): 730–737.
- **Di Tella, Rafael, and Ernesto Schargrodsky.** 2004. “Do Police Reduce Crime? Estimate Using the
Allocation of Police Forces after a Terrorist Attack.” American Economic Review, 94(1): 115–133.
- **Dur, Robert, and Ben Vollaard. 2019.** “Salience of law enforcement: A field experiment.” Journal
of Environmental Economics and Management, 93: 208–220.
- **Cheng, Cheng, and Wei Long. 2018.** “Improving police services: Evidence from the French quarter
task force.” Journal of Public Economics, 164: 1–18.

---
class: segue-yellow
background-image: url("assets/TulaneLogo.svg")
background-size: 20%
background-position: 95% 95%

# Terminology

---
## Some Terms that are Helpful to Know

- These terms will come up in the practice questions you’ll do today, in the course, and in economics in general.
1. Treatment variation
1. Exogenous
1. Endogenous
1. External validity

---
## Treatment Variation

- This term often comes up in empirical research that estimates causal effects.
- Treatment variation refers to the variation in X that you are using to identify the causal effect of X on Y.
- E.g., the variation in the timing and location of MDPs (as in GHM).
- E.g., the increase in police officer hiring that occurs before elections (this is the variation used in Levitt, 1997).
- E.g., the randomly-assigned change in police enforcement by location (as in Dur and Voolaard, 2019).

---
## Exogenous

- If a variation is exogenous, then it is not a function of other factors.
- It is not a function of other variable in the economic/statistical model.
- If something is exogenous, you can think of it being random.
- We ideally want to use treatment variation that is exogenous. 
- The gold standard would be randomization $\rightarrow$ randomized treatment (e.g., randomizing police) would be strictly exogenous since treatment status does not depend on anything.

---
## Endogenous

- The opposite of exogenous.
- More specifically, if something is endogenous, it means it is a function of (it depends on, it is endogenous to) something else.
- E.g., police allocation is endogenous to crime (unless we find some random or quasi-random variation to use).
- E.g., state adoption of tax incentives for the film industry may be endogenous to the size of the existing film industry (larger existing film industry = more likely to adopt an incentive)

---
## Endogenous vs. Exogenous

- The key question is how exogenous/endogenous the treatment variation is. 
- It’s not black and white where it is always clearly one or the other.
- Most treatment variation outside of an experiment lies on a spectrum between fully exogenous and strongly endogenous.
- There is no way to know or to test of treatment variation is endogenous.
- Determining how endogenous it is requires thinking critically about the factors that affect the treatment variation.
- For example, is there something non-random about the change in policing that is used in the paper? Could this non-randomness cause bias by creating a feedback loop (like the crime $\leftrightarrow$ police feedback loop shown earlier)?


```{r gen_pdf, include = FALSE, cache = FALSE, eval = TRUE}
library(renderthis)
# https://hhadah.github.io/xaringan/index.html

to_pdf(from = "~/Documents/GiT/urban-econ/slides/06-week6/01-tue/causal-term.html", 
       to = "~/Documents/GiT/urban-econ/slides/06-week6/01-tue/causal-term.pdf")
to_pdf(from = "~/Documents/GiT/urban-econ/slides/06-week6/01-tue/causal-term.html")
```
